\subsection{S3K}
S3K is a real-time operating system that provides a capability-based multicore
partitioning kernel for embedded RISC-V systems. It provides spatial and
temporal isolation\footnote{Spatial isolation means separation of memory,
whereas temporal isolation means that the failure of one component should not
affect the schedulability of another.}, time protection, and dynamic resource
reconfiguration. Furthermore, S3K's scheduler guarantees deterministic process
dispatch, free from microarchitectural interference. \cite{karlsson2025}

It achieves this through three main attributes:
\begin{itemize}
  \item A capability model.
  \item A time-driven, capability-based scheduler with systematic use of
    \texttt{fence.t}, worst-time padding, and constant-time kernel operations.
  \item Predictable, constant-time system calls.
\end{itemize}

\subsubsection{Implementation}
The implementation provides a minimal trusted computing base (TCB) with strong
isolation and security guarantees. The basic execution unit in S3K is a
\emph{process}, a self-contained entity whose resources are defined by its
capabilities. These processes are designed to host bare-metal real-time
operating systems or applications, with the kernel acting as a bare-metal
hypervisor. The number of processes is statically defined at compile time.
Processes use \emph{monitor capabilities} and capability revocation to manage
other processes, reclaim resources, and repurpose existing processes for new
tasks. \cite{karlsson2025}

\paragraph{Capabilities:}
Every S3K capability operation has a bounded worst-case execution time (WCET)
and is free of timing side-channels. Capabilities are implemented in software
and stored in a capability derivation tree (CDT), which resides in a statically
allocated capability table (ctable) in memory.

\paragraph{Scheduling:}
S3K's scheduler is a partitioned scheduler that divides time into \emph{major
frames}, which are further partitioned into \emph{minor frames}. S3K treats
minor frames as first-class capabilities, enabling dynamic reconfiguration of
computing resources while maintaining process isolation. \cite{karlsson2025}

A major frame is divided into a fixed number of equal duration time slots. A
single time slot represents the smallest scheduling unit. A time slice
capability grants control over a range of time slots on a hardware thread
(hart). The process that owns the time slice capability receives a minor frame
of its own, whose length is less than or equal to the major frame. Each time
slice capability has an \emph{enable} field that specifies whether the minor
frame is active. If set, the owner of the time slice is scheduled during the
corresponding minor frame; otherwise the minor frame represents idle time.

The kernel is in principle non-preemptable. It uses preemption points to ensure
that the WCET is bounded. The kernel checks for preemption at the start of
every system call and aborts the system call if preemption is required.

By the definition of domains in S3K, processes from different domains are never
scheduled in parallel. Thus, during a domain switch, all running processes of
the current domain are preempted, invoke the scheduler, and are subsequently
delayed by the temporal fence.

\subsubsection{Time protection}
S3K uses the temporal fence instruction at context switches to flush the
core-local microarchitectural state.

S3K is designed to reside entirely within the small scratchpad memory (SPM)
backed by the core-local L1 cache. This enables the kernel's microarchitectural
footprint to be efficiently flushed and avoids reliance on larger caches.

S3K only protects a process's in-kernel state from side-channels; it does not
protect a process's user-level state. Each process is responsible for protecting
its own user-level execution from side-channels in higher level caches. This is
different from the approach in \cite{ge2019}, where cache coloring guarantees
that processes in different domains never access the same parts of the cache,
thereby preventing side channels even in larger caches.

All system calls are designed to be non-interfering and constant-time with
respect to the invoking process's authorized observations; that is, all control
flow decisions and memory accesses depend only on data the process is permitted
to observe.

\subsubsection{Evaluation}
They evaluate the implementation across several factors, the most significant
for our work being scheduling jitter and side-channels. They first find that
Inter-Process Communication (IPC) operations have the highest recorded
non-preemptable cost of 1,195 CPU cycles. Together with an estimated 16,000
cycles for data cache write-back during the temporal fence, they set the CSPAD
to 18,000 cycles for the \texttt{fence.t} instruction. The make use of these
findings for the WCET padding throughout the codebase.

\paragraph{Scheduling Jitter:} A measurer process, scheduled once per major
frame, records the \emph{mcycle} performance counter value upon dispatch. To
simulate interference, they introduce interfering processes that execute random
system calls and pollute the cache. They find that when the measurer exclusively
occupies the SPM, the other processes have no impact on the measurements.
However, when the measurer occupies DRAM, noticeable jitter appears. They note
that in some cases this jitter originates from the small portion of code in the
measurer that records the dispatch time, rather than from the scheduler itself.
When more than one process is scheduled at a time, jitter also appears in the
scheduler.

\paragraph{Side channels:} They implement a Trojan process and a spy process.
The Trojan signals a "0" by remaining idle and a "1" by behaving like the
interfering processes from before. The spy then measures jitter, system call
execution times, and cache access times to infer the Trojan's signal.

They find that when the measurer and Trojan both reside in DRAM, there is a
near perfect information flow. In contrast, when both reside in the SPM, there
is no measurable amount of information flow.
