\subsection{temporal\_fence\_t}\label{sub:temporal} % (fold)
Following the difficulties found in \cite{ge2019}, it was clear that hardware
needs improvements before time protection truly could be achieved. In the paper
\cite{wistoff2023} they implement such a hardware improvement. First, it is
demonstrated that there currently exists presence of serious timing channels
within the in-order RISC-V core and confirm that the current ISA lacks
mechanisms to close them reliably.

The initially present two versions of the \texttt{fence.t} instruction, a basic
flush which exclusively flushes the L1 data and instruction caches, the TLBs,
and the branch predictors. Through their evaluation they find that further
stateful components such as the LFSR providing pseudo-random index sequence for
the cache replacement policy and the round-robin memory arbiters of the core
still leak data in the microarchitectural state. As such, they also implement a
Full Flush, which clears these components along with the ones in the basic
flush. To use this full flush efficiently the \texttt{fence.t} proceeds as follows.
\begin{itemize}
  \item Save the program counter.
  \item Save locally modified (dirty) architectural state.
  \item Drain pending transactions.
  \item Clear components that are not cleared on reset.
  \item Assert Microreset. Clearing all flip-flops containing non-architectural state.
\end{itemize}

A full explanation of each step is provided in \cite{wistoff2023}.

Given this implementation, they augmented the seL4 kernel to always call the
\texttt{fence.t} instruction whenever there is a context switch, and they find
that the overhead is around 0.7\%. It is noted, that this should be seen as a
pessimistic result due to the use of idle tasks, and that in a setting with
high memory contention, the latency may be almost entirely hidden.

\subsubsection{With software support}
While \texttt{fence.t} was effective and inexpensive on simple in-order cores.
However, in a follow up paper \cite{wistoff2025} argue that in out-of-order
execution CPUs there are challenges with the implementation and address these
challenges by presenting a SW-supported temporal fence \texttt{fence.t.s}
methodology. They show that this software supported version is able to close
all on-core timing channels without measurable hardware overhead at a minimal
performance impact of 1.0\% on average.


% section temporal\_fence\_t (end)
