\subsection{User-level mitigations}
The argument for using user level mitigations stems from the definition of time
protection. Time protection is defined within \cite{ge2019} as "A collection of
OS mechanisms which jointly prevent interference between security domains that
would make execution speed in one domain dependent on the activities of
another". An OS, which implements Time protection must ensure that all
execution within partitioned parts does not depend on each other making both
covert channels and side-channels infeasible. This is quite a strict version of
time protection, and to close covert channels it must be, but quite a
significant number of timing channels can be closed at the user-level if one
allows a distinction between different information types.

One such example is the FaCT DSL \cite{cauligi2019}. FaCT sought to address the
challenges of writing constant-time cryptographic code, which does not depend
on "secret" information. That is, instead of treating all information as
possible leakage, one indicates what information must not be leaked through a
timing channel, and the FaCT type system guarantees that in isolated execution
this will not leak. FaCT works by making the code constant time in regard to
the sensitive information, and any branching is only dependent on non-sensitive
information. The idea being that any attack can not measure the execution time
of the encryption code to gather information about sensitive information. This
does work to an extent, by guaranteeing constant time at the C program level,
but compiler-introduced and variable execution times of instructions based on
inputs can still leak sensitive information. For example, the division operator
is notoriously known to have its execution dependent on the numerator
\cite{schroder, jin2023}. Mitigations against these attacks would still require
help from hardware, and make use of specific instructions that guarantee a
constant execution time.

Furthermore, information that the operating system holds about a given process
could also be deemed sensitive. For example, in highly secure environments the
runnability of a task could be deemed sensitive, and no changes on the user
level side will ever close such timing channels, as the OS scheduler will
unknowingly always leak this information through its scheduling decisions.
Therefore, while user level implementations do help significantly, the issue
can not be fully closed without both the support of the OS and hardware.

So while user level mitigations are effective at closing a significant amount
of timing channels, it is not capable of providing complete time protection by
itself, and still requires help from the OS and hardware to do so. The user
level mitigations are thus not useable in the context of full time protection,
but could be used in cases where restrictions are more lenient.
