\section{Background}\label{sec:Background}
To address the challenges of time protection, it is essential to first
establish a general understanding of how these vulnerabilities manifest and key
terms used when explaining timing channels. This section provides the technical
foundation for the work that follows and aims to build a foundational
understanding about what channels are and how isolation can be used as a means
for protection.

First, Section~\ref{sub:side-channel} explores side-channels, detailing how
tasks can involuntarily leak information through execution time and
speculative execution. Section~\ref{sub:covert-channel} investigates the
extension of this, where two colluding processes intentionally attempt to
bypass security boundaries. Section~\ref{sub:timing_channels} categorizes the
specific timing channels an operating system faces, splitting them into
scheduler-based, shared-state and delay-based vulnerabilities.
Section~\ref{sub:isolation} shifts the focus toward mitigation, detailing the
principles of spatial and temporal isolation, before
Section~\ref{sub:temporal} reviews the hardware-assisted \texttt{fence.t}
instruction designed to eliminate microarchitectural state leakage with
temporal isolation. Finally, Section~\ref{sub:FreeRTOS} introduces the
FreeRTOS ecosystem and the RISC-V platform used as the foundational
environment for this thesis.

\subsection{Side-channels}\label{sub:side-channel}
Side-channels are unintended information channels that leak information about
the process. These channels infer secret data often through observable
execution time of a process. The general case of such a channel can be seen
in the snippet below:

\begin{algorithm}[H]
  \floatname{algorithm}{Naive side-channel example}
  \begin{algorithmic}[1]
    \If{secret}
      \State{Some long computation}
    \Else
     \State{Do nothing}
    \EndIf
  \end{algorithmic}
\end{algorithm}

In this example, the program branches on a secret value, and the differing
execution times of the two paths leak information about the secret. If, for
instance, this algorithm were part of a library, an attacker could infer the
secret by measuring the execution time of the call. This type of side-channel
can be prevented through constant-time programming at the user level, where
both branches perform an equally long computation. However, this approach
remains fragile: even when the source code appears constant-time, compilers may
introduce optimisations that reintroduce timing side-channels, as shown in
\cite{schneider2025}.

Further, more complex examples of side-channels includes those specified within
\cite{kocher2019}, where speculative execution can be used to leak information
about an otherwise safe algorithm. An example from the paper is the following:

\begin{algorithm}[H]
  \floatname{algorithm}{Spectre side-channel example}
  \algrenewcommand\algorithmicrequire{\textbf{Input: some unsigned integer, x}}
  \begin{algorithmic}[1]
    \If{$x < array1\_size$}
    \State{$y = array2[array1[x] * 4096]$}
    \EndIf
  \end{algorithmic}
\end{algorithm}

This function could again be part of some library or system call, which
receives an unsigned integer x from the user. The process, which runs this
library code or system call has access to both array1 and array2. It does a
bounds check on x, to ensure there is no out of bounds access, which could
trigger an exception or access sensitive memory. If this algorithm was called
multiple times, with correct input, i.e. where $x < array1\_size$, then an
attacker could train the branch predictor on modern CPUs making the CPU
speculatively execute the expected path, which could bring in the location read
from \texttt{array2} into the cache, and the attacker could then use attacks
such as Flush+Reload or Prime+Probe to reveal which part of array2 was loaded
into the cache. With this, it can gather the index of the loaded item, and can
figure out what was read from array1[x], which could be some arbitrary secret
obtained through the speculative out-of-bounds access enabled by the branch
predictor. Such secrets might be sensitive data residing in memory immediately
after the \texttt{array1} buffer, for instance cryptographic keys or passwords.
A more detailed explanation along with a proof-of-concept is provided in
\cite{kocher2019}, but this small example shows how seemingly safe code, might
still leak secret information.

A multitude of such attacks, which make use of these side-channels are
presented in \cite{osvik2006, liu2015}.

\subsection{Covert channels}\label{sub:covert-channel}
While side-channels are passive and exploit unintended leakage, covert channels
involve two processes working together.  This usually revolves around a sender
(Trojan) and a receiver (spy), where the sender has access to high security
data but is restricted in its communication with the receiver. To bypass this,
the sender and receiver collude by any means possible to share information.
This could be via the same techniques as the side-channels, but it could also
be any use of shared state between the two processes.

Generally two main categories of covert channels exist; \textbf{Storage
channels:} which modify some stored object, e.g. a file or some metadata;
\textbf{Timing channels:} which change the time of different observable events
for the receiver. For this thesis storage channels are not referenced, as
kernels such as seL4 have already formally proven the absence of them
\cite{murray2013}.

Covert timing channels work through the same mechanisms as the aforementioned
side-channel, but pose as a worst case, where the sender can make use of any
and all leakage methods to convey the information. This way, one is not limited
to the visible gadgets that might exist in a vulnerable program, but the
information might also be conveyed through repeated use of non-constant time
system calls that depend on some shared resource, or as the aforementioned
example shared use of the cache.

\subsection{Timing channels}\label{sub:timing_channels}
Both side-channels and covert channels rely on some shared resource whose state
or timing can be observed by a receiver. In the context of an operating system
kernel, these shared resources take several distinct forms. This subsection
presents some of the timing channels an OS kernel could face, categorising the
items that enable the channel.

\paragraph{Scheduler-based channels} The OS scheduler itself can function as an
unintended information channel. The scheduler controls which tasks get to run
and might inadvertently leak important information about the schedulability of
a task through this behaviour. Examples of a side-channel in the priority based
real time schedulers can be seen in \cite{chen2019}, where they demonstrate how
an unprivileged, low-priority task (in user space) can infer execution
behaviours of critical, high-priority periodic tasks. They show how this
information is useful to launch other more devastating attacks and demonstrate
two such examples. Further, in \cite{kadloor2016} they study more general
shared schedulers and how information leakage arise in the context of shared
event schedulers and argue about the different trade-off between privacy and
performance.

A simple example of a scheduler-based channel could be one present in a
priority-based scheduler. In such a scheduler the highest priority task can
inhibit execution of lower priority tasks. This could be used as a covert
channel, where the higher priority task can signal either a 0 or 1 by simply
going to sleep, thus allowing the lower priority task to run, or it could block
execution of the lower priority task. The lower priority task can measure its
last wake time, and infer if the higher priority task attempted to signal
either a 0 or 1.

\paragraph{Shared state channels}
Microarchitectural resources such as caches, TLBs, and branch predictors are
typically shared across tasks executing on a core. As demonstrated in
Section~\ref{sub:side-channel}, the state left in these components by one task
can be observed by another through timing measurements, creating a class of
channels that is both broad and difficult to eliminate in software alone. These
are the channels that often require explicit help from hardware mechanisms such
as the temporal fence instruction \cite{wistoff2023, wistoff2025}, which are
investigated further in Section~\ref{sub:temporal}.

\paragraph{Delay-based channels}
The interrupt mechanism itself, together with the latency it introduces, can
carry timing information. A task could trigger different interrupts, each
delaying another task's execution and thereby forming a timing channel. One
might also state that external interrupts could be sensitive information, which
again would be a source of timing channel depending on how the OS handles this.
In a related manner, when a task holds a lock it prevents another task from
acquiring the same resource, which again delays execution and leaks information
between the two tasks.

Mitigating these channels requires a combination of mechanisms to provide true
time protection.

\subsection{Isolation}\label{sub:isolation}
Throughout many timing channel mitigations, isolation is a common factor.
Generally there exist two types of isolation, namely \textbf{spatial} and
\textbf{temporal} isolation. Spatial isolation partitions software into
distinct memory spaces, which prevents separate domains from sharing the same
underlying hardware. This separation is most commonly used for DRAM memory,
where the OS kernel relies on hardware support like the memory management units
(MMUs) or memory protection units (MPUs). The MMU translates virtual memory
addresses into physical addresses in main memory and the OS can separate
different processes usage entirely. It also generally provides memory
protection by implementing access control rules and regulations stopping
illegal usage of particular memory locations. 

The latter isolation type, namely temporal isolation, must ensure that the
service received from shared resources by the software in one domain cannot be
affected by the software in another domain \cite{rushby1999}. An example, which
would require such isolation is any shared part of the OS between different
domains. Calling shared OS code should not have any measurable timing effect
on another domain calling the same OS code. The solution is often to reset the
shared resource back to a known default before any partitioned item is run.
This guarantees that each process runs from a clean state which is unaffected
by previous use.

To achieve temporal isolation one must clean any microarchitectural state that
could be affected by process usage. This is where an instruction such as
\texttt{fence.t} \cite{wistoff2023} would come into play. The instruction
clears any shared microarchitectural state that cannot be partitioned spatially
allowing for a deterministic state for a new process to run in.

\subsection{The \texttt{fence.t} instruction}\label{sub:temporal} % (fold)
Following the difficulties identified in \cite{ge2019}, it was clear that
hardware improvements are needed before time protection can be fully achieved.
The paper \cite{wistoff2023} introduce one such hardware improvement. First, it
demonstrates the presence of serious timing channels within an in-order RISC-V
core and confirms that the current Instruction Set Architecture(ISA) lacks
mechanisms to close them reliably.

Following this, they present two versions of the \texttt{fence.t} instruction:
a basic flush that exclusively flushes the L1 data and instruction caches, the
TLBs, and the branch predictors. Through their evaluation they find that
further stateful components - such as the LFSR generating a pseudo-random index
sequence for the cache replacement policy and the round-robin memory arbiters
of the core - still leak data through the microarchitectural state. As such,
they also implement a Full Flush, which clears these components along with
those covered by the basic flush. The \texttt{fence.t} instruction proceeds as
follows:
\begin{itemize}
  \item Save the program counter.
  \item Save locally modified (dirty) architectural state.
  \item Drain pending transactions.
  \item Clear components that are not cleared on reset.
  \item Assert Microreset. Clearing all flip-flops containing non-architectural
    state.
\end{itemize}

A full explanation of each step is provided in \cite{wistoff2023}.

Given this implementation, they augmented the seL4 kernel to call the
\texttt{fence.t} instruction on every context switch, finding an overhead of
around 0.7\%. They note that this should be seen as a pessimistic result due to
the use of idle tasks, and that in a setting with high memory contention the
latency may be almost entirely hidden.

\subsubsection{With software support}
While \texttt{fence.t} was effective and inexpensive on simple in-order cores,
a follow-up paper \cite{wistoff2025} argues that out-of-order execution CPUs
introduce challenges for the implementation. They address these challenges by
presenting a software-supported temporal fence \texttt{fence.t.s} methodology.
They show that this software-supported version is able to close all on-core
timing channels without measurable hardware overhead, and with a minimal
average performance impact of 1.0\%.

% section temporal\_fence\_t (end)

\subsection{FreeRTOS}\label{sub:FreeRTOS} % (fold)
FreeRTOS is a class of real-time operating systems (RTOS) designed to be small
enough for microcontrollers, while not limiting its application to
microcontrollers alone. Out of the box, FreeRTOS provides core real-time
scheduling functionality, inter-task communication, timing and synchronisation
primitives, along with other optional add-on features
\cite{WhatFreeRTOSFreeRTOS2026}. It does this through the use of a
priority-based scheduler, which is inherently not fair and can starve lower
priority tasks.

FreeRTOS-MPU is a memory-protected version of FreeRTOS, which at the time of
writing is supported on select ARM-based microcontrollers. It enables tasks to
run in either privileged or unprivileged mode and restricts access to resources
such as RAM, executable code, peripherals, and memory beyond the limit of a
task's stack \cite{MemoryProtectionUnit2026}.

A third-party port of FreeRTOS-MPU to the RISC-V ISA currently exists
\cite{FreeRTOSCommunitySupportedDemosRISCV_RV32_MPU_QEMU_VIRT_GCCmain}. This
port forms the basis for this thesis's implementation. There are several reasons
for this choice. The first and most important is familiarity: I have previously
worked with the RISC-V ISA, which reduces the overhead of adopting a different
ISA. Second, the related work on timing channels generally focuses on RISC-V, so
future comparisons between implementations are more relevant if all are built on
the same platform. Finally, the \texttt{fence.t} instruction introduced in
Section~\ref{sub:temporal} is designed with the RISC-V instruction set in mind.
While the underlying idea could be implemented on other platforms, work on
hardware support for those platforms is not as far along. Therefore, assuming
this implementation will eventually run on real hardware, RISC-V seems the most
reasonable choice.

% subsection FreeRTOS (end)

