\section{Conclusion}
It is clear that timing channels pose a security threat in modern operating
systems. As more of the low hanging fruit gets patched, these advanced methods
of breaking intended OS barriers become more prevalent. The FreeRTOS default
priority-based scheduler is inherently vulnerable to these types of timing
channels, and can be used to leak information via timing channels by two
collaborating tasks. Prior work exists in creating secure microkernels
\cite{ge2019, karlsson2025} as well as promising developments in the user-level
mitigations \cite{cauligi2019}. However, there still exists a gap in how time
protection can be implemented in a reasonable manner in existing operating
systems.

This thesis attempts to bridge this gap, by implementing a domain-level
scheduler within the existing framework of the FreeRTOS kernel. It does so,
while keeping true to the foundation of FreeRTOS and keeping the priority-based
scheduler within each of these domains. On a given domain switch, the domain
level scheduler makes use of the temporal fence instruction \cite{wistoff2023,
wistoff2025} to clear the microarchitectural state eliminating. Another focus
was placed on the domain creation to be constant-time, so as not to
inadvertently leak information. This was all implemented as a modification of
the FreeRTOS RISC-V MPU port, such that the MPU protections can be used to
completely separate domains. Further, an intended communication channel was
added between domains by modifying the Queue message passing native to
FreeRTOS, so that it functions both cross-domain and within domain.

The evaluation shows that the implementation is functional, and that in regards
to the QEMU specific system emulation it has constant-time time slices. The
round trip times have a cycle count of 50000 with a variance of just 1 cycle.

The implementation is not without its limitations. Notably it changes some of
the fundamental expectations in FreeRTOS. As an example the vTaskDelayUntil
function is only relative to within a domain as it depends on a within-domain
xTicks that is only incremented when the domain is run. Due to the
constraints of the current MPU port, any privileged task can break all
partitioning and access data across domains without restrictions, meaning that
only user mode to user mode cross-domain protection is enforced. Due to the
large code base, not all of the system calls are checked for information
leakage, and as such it could be possible there exist unknown channels here.
Lastly, higher-level caches are not cleared with the temporal fence instruction
and remain vulnerable. This includes the kernel memory usage itself, as the
implementation does not make use of a scratchpad memory as seen in the S3K
kernel \cite{karlsson2025}.

Further, the evaluation itself was significantly limited by the lack of real
world hardware. The evaluation does not take into account the underlying
microarchitectural state and as such the implementation focused solely on the
scheduling based timing channels. While it does make use of the temporal fence
instruction on domain switching, this functionality remains untested given the
testing constraints.

Even with this, this thesis has contributed by showing that time protection can
be added into an existing mainstream kernel, while showing intended behaviour
on the QEMU system emulation platform. It shows that time protection is
achievable, but comes with significant trade-offs in complexity, performance
and usability. So while it is possible, a full domain-level scheduler is most
likely not suitable for more generic operating systems, but would remain viable
for environments where a specialized microkernel with security is needed. 

With this in mind, there are still aspects of time protection that could be
used in the more generic operating systems. Specifically, future research can
make use of these concepts:

\begin{itemize}
  \item \textbf{Selective Temporal Partitioning: }Given the low performance
    overhead of the temporal fence instruction (1\% on average), it could be
    used in a generic operating system to create sensitive tasks. The operating
    system could flush the microarchitectural state before and after these
    sensitive tasks are scheduled. This would eliminate their
    microarchitectural footprint while allowing for more freedom in execution
    for the sensitive tasks.
  \item \textbf{Hardware-Assisted partitioning: } In conjunction to this,
    hardware features such as cache partitioning and scratchpad memory could
    guarantee that non-flushable components also remain partitioned from these
    specific sensitive tasks.
  \item \textbf{User level cooperation: } With safe hardware, the user level
    mitigations could cover the remaining attack front in these general purpose
    environments, and could lead to a safe execution environment for known
    sensitive tasks.
\end{itemize}

However, to fully guarantee this security contract, significant future work
remains to formally prove time protection as a whole, while also validating the
correctness for this specific use case.
