\section{Description}
\subsection{Background and Motivation}
For decades, the dominant threat model for software security has centered on
memory safety vulnerabilities, including buffer overflows, format-string bugs,
use after free errors, and integer conversion flaws. These vulnerabilities have
historically been the industry’s “low hanging fruit” for attackers, because
exploitation is straightforward and consequences are severe. Modern
cyberattacks still rely heavily on these vulnerabilities, which remain in
systems software written in C and C++.

In response, the industry is increasingly adopting strong countermeasures,
including memory-safe languages like Rust, sophisticated compiler mitigations
(e.g., stack canaries), and dynamic analysis tools. As these defenses mature
and memory exploits become harder to execute, attackers are shifting their
attention to "higher-hanging fruit" including but not limited to: timing
channels, cache-based side channels and interrupt-based leaks all part of the
term "side-channel attacks".

Unlike memory corruption, which alters the program's state or control flow,
timing attacks exploit the physical execution of the software. Timing channels
arise when execution time depends on secret or security-relevant internal
states. An attacker can observe execution time, and infer sensitive information
such as private keys, memory-access patterns, scheduling decisions, or resource
contention. Timing channels violate intended isolation boundaries at the
process, VM, or network level, allowing seemingly independent programs to leak
information indirectly. Such attacks have already been shown to be detrimental
and as such it is not something that can be left untouched \cite{spectre,
meltdown}.

A variety of countermeasures have been proposed to address these issues. In
cryptography domain specific languages such as FaCT provide transformations of
potentially timing sensitive high level code to low level constant time code
\cite{cauligiFaCTDSLTimingsensitive2019}. This ensures that in an isolated
environment the low level code will not leak information about its secrets
through timing side channels. However, practical systems rarely operate in such
controlled environments: operating systems provide no guarantees regarding
scheduling decisions and do not flush caches or registers on context switches.
Consequently, opportunities for microarchitectural channel leaks can still
occur. 

Previous work has been done to attempt to prevent such leaks. For instance
hardware specific extensions have been implemented which provide mechanisms for
clearing vulnerable micro architectural state and guaranteeing a
history independent context switch latency \cite{fencet}. Other work has
examined the work required for time protection and have provided an
implementation for the seL4 microkernel showing it is possible to protect
against with little impedance on performance \cite{MissingOSAbstraction}.

However, evaluating timing-channel resistance remains difficult. Existing
approaches often rely on empirical measurements on real hardware, specialized
platforms such as FPGAs, or manual inspection of execution traces. There is
currently no widely available, systematic platform for analyzing and reasoning
about cache-induced timing channels in a controlled and reproducible manner.

This thesis addresses this gap by focusing on the cache layer as a primary
source of timing leakage and by providing an emulation-based framework for
analyzing and mitigating such channels.

\subsection{Problem statement}
Despite their importance, timing side channels, particularly cache based
channels, remain insufficiently defended in mainstream systems. While
constant time programming techniques can eliminate data dependent control flow
at the instruction level, they do not prevent leakage caused by shared cache
state, context switching, or operating system scheduling decisions.

The central problem this thesis addresses is: \emph{To what extent can timing
side-channels be mitigated through a combination of hardware aware software
design and what platform is needed to prove timing channel mitigations?}

Specifically, this project will provide a plugin integration to the QEMU
virtual machine. This implementation models cache behavior for configurable
cache architectures and enables detection of potential timing channels arising
from cache contention, eviction patterns, and shared microarchitectural state.
The goal is to provide a deterministic and reproducible environment in which
timing-channel behavior can be studied without relying solely on empirical
hardware measurements.

This implementation will be tested using previous "timing" safe microkernels,
such as the \cite{S3K}, and will further analysis the behaviour of a
microkernel, which does not take timing channels into account. Further, this
project aims to augment the "unsafe" microkernel using varies flushing
strategies, constant time programming and partitioning to evaluate how
operating system design can reduce cache based timing leakage.

This project further wants to establish hardware requirements for such
mitigations to work efficiently and what division of responsibility between
hardware and software is required to enable practical timing-secure execution
in microkernel based systems.

\subsection{Project-Specific Learning Objectives}
\begin{itemize}
  \item Explain and understand the theoretical and practical mechanisms of
    side-channel leaks across hardware, OS and program layers.
  \item Survey existing side-channel mitigation techniques including but not
    limited to: hardware mechanisms and OS-level strategies.
  \item Design and implement a cache emulation layer capable of finding timing channels in the cache layer.
  \item Evaluate the efficacy and viability of the implementation by running already established microkernels.
  \item Modify existing microkernel to provide better timing channel
    mitigations and evaluate the efficacy along with performance hindrance.
  \item Gather the findings into design recommendations for future operating
    systems, hardware abstractions or tool chains, showing how side channel-secure
    execution could be supported in the future. 
\end{itemize}

