\section{Future work and discussion}\label{sec:FutureWork}
While the implementation does show promise, there are still significant items
that are left for future work. This section addresses these remaining
challenges across four main areas. First, Section~\ref{sub:cycle-accurate}
discusses the transition from virtualised environments to cycle-accurate
hardware emulation. Section~\ref{sub:safeguard-syscalls} then analyzes the
challenges of safeguarding system calls and how to prevent them from
inadvertently becoming a timing channel. Next, Section~\ref{sub:FreeRTOS-vast}
outlines the necessity of a comprehensive audit of the full FreeRTOS kernel
before any guarantees of time protection, as the shared parts of the kernel
could still pose issues. Finally, Section~\ref{sub:remaining-channels} explores
the remaining known timing channels, which are not fixed by scheduler design or
flushing the microarchitectural state.

\subsection{Verification on cycle-accurate hardware}\label{sub:cycle-accurate}
As mentioned previously, the QEMU virtual machine does not provide cycle
accurate emulation. The closest provided is the instruction-based time counting
as described in \cite{TCGInstructionCounting}. This means that any timing
channel that arises from the microarchitectural shared state cannot be
emulated or diagnosed on the QEMU platform. What can be empirically evaluated
on the platform is a constant time execution path as shown with the previous
domain-level scheduler. 

During the evaluation phase of this thesis a QEMU TCG plugin was developed in
an attempt to bridge this gap. The plugin provided a cache emulation of the L1
and L2 cache with support for flushing the L1 cache on domain switches, and it
functioned correctly in emulation. However, QEMU was never designed to produce
cycle-accurate emulation, and when the plugin attempted to update the machine
timer externally, this happened asynchronously and was affected by the host
machine running QEMU. This meant that even though the cache emulation seemed
correct, there would be a non-deterministic execution from the guest OS point
of view, making time control unusable. While significant work went into the
plugin, it ultimately proved unable to provide the cycle-accurate emulation
necessary for timing channel analysis. To properly evaluate the implementation,
a hardware platform that supports the temporal fence instruction would be
ideal. As an alternative, cycle-accurate simulators such as gem5
\cite{lowe-powerGem5SimulatorVersion2020} could prove valuable for future
evaluation. While other research has identified some challenges with gem5
\cite{bossuet2023}, such a simulator could significantly reduce the barrier of
entry and lead to more rapid development in the pursuit of time protection.

\subsection{Safeguarding system calls}\label{sub:safeguard-syscalls}
With the current setup, when a new domain is created, it must be created with a
task that is in privileged mode, if that domain should have more than a single
task running. There are two primary reasons for this. First, to create a task
in FreeRTOS-MPU one must be in privileged mode. Secondly, to add a task to a
domain with the new API presented in Section~\ref{sub:api}, one must either be
inside the domain itself or be creating a new domain\footnote{Given the task
  runs in privileged mode, it could technically overwrite the memory locations
forcefully, and not make use of the api at all.}. The requirement of a single
privileged task mainly comes from the FreeRTOS-MPU version, and to circumvent
this issue would require significant changes to FreeRTOS itself. The larger
issue in regard to domain creation is that creating a new domain could delay
scheduling. Specifically, many of the privileged functions enter a "critical"
section within FreeRTOS. When entering such a critical section, on the RISC-V
MPU version interrupts are disabled, including the machine timer interrupt,
which is the foundation of the domain scheduling. This poses a significant
burden on the privileged tasks within each domain, as they can inadvertently
create timing side-channels. The issue is not only with the domain creation,
but any call to functions that enter the critical sections. 

This issue poses a few challenges, and there are two primary solutions, which
could be considered for future development. The first is taking inspiration
from \cite{ge2019}. Here, they clone the shared kernel data into memory
provided by user level tasks within each domain. The current implementation
created within FreeRTOS does not do this, but many of the preexisting features
of the within domain priority scheduler could be fully partitioned. For this to
function, one would have to introduce two different kinds of critical sections.
There is the cross-domain critical section, for which it would still be
necessary to disable interrupts, and there is the within domain critical
section, which only runs the priority-based scheduler. When entering a within
domain critical section, then the priority-based scheduling would block
scheduling, such that the priority-based scheduler is not left in an invalid
state. However, if there is a domain-level interrupt, then the context would be
saved of the priority-based scheduler, and a switch to another domain would
still happen. When the old domain runs again, it would still be within the
domain critical section at the same context point, allowing for a valid state
of the priority-based scheduler, while allowing the domain-level scheduler to
schedule other domains without delays. This context saving and restoration
would have to be taken into account for the worst case analysis of the domain
switching, if it can not be guaranteed in constant time, but should propagate
well to the majority of FreeRTOS' standard implementation.

The next issue would be the cross-domain critical section. This part of the
kernel is shared between all domains, and as such one cannot simply "ignore"
the critical section as before. Instead one would have to guarantee that any
call to cross domain critical functions would finish execution before the
current domain is interrupted. A solution to this problem would be to guarantee
that any system call with such a cross-domain critical section runs within a
single time slot, or allows for preemption within a time slice. Then, when
entering the cross-domain critical section, the task would have to idle until
it reaches the start of a time slice and then execute the system call. This
would mean that execution is guaranteed to finish before the scheduler would
attempt to schedule the next domain. 

For this to function correctly one would need proper hardware to find the
running time of all such system calls, and make sure that the execution time is
bounded by the time slice. It would also mean that there has to be a hard
requirement on the lowest possible value of the time slice duration. However,
with these in mind truly isolated domains should be possible.

Note however, that this side-channel would only be possible during the end of a
domain's time slice, and would only be able to pass information between the
current domain and the domain which comes immediately after. As an example, I
believe this could be a possible timing channel within the S3K kernel. S3K
makes use of a global TTAS lock. A lock is acquired through the following code
snippet:

\begin{lstlisting}[language=c]
bool ttas_acquire(ttas_t *ttas, bool preemptable)
{
	while (__atomic_exchange_n(&ttas->lock, 1, __ATOMIC_ACQUIRE)) {
		if (preemptable && preempt()) {
			return false;
		}
	}
	return true;
}
\end{lstlisting}

The preemptable flag is used to indicate whether the current lock acquisition
will respect a preemption. The preempt function call checks if the mip register
has a bit set at (1 << 7), which corresponds to the Machine Timer Interrupt
Pending (MTIP) bit \cite{MipRISCVISA}. This bit indicates that the machine
timer is pending and should fire. In S3K this lock is acquired in two important
places, namely:

\begin{lstlisting}[language=c]
proc_t *syscall_handler(void)
{
	// The system call number.
	word_t syscall_nr = current->regs.a0;

	// If system call number is invalid, make an exception.
	if (syscall_nr >= ARRAY_SIZE(handlers)) {
		return exception_handler(0x8, syscall_nr);
	}

	// Try to acquire a lock. Also checks for preemption.
	if (!lock_acquire(true)) {
		return NULL;
	}

	// Advance the program counter.
	current->regs.pc += 4;

	// Call the system call handler
	proc_t *next = handlers[syscall_nr](current->pid, &current->regs.a0);

	// Releases the lock.
	lock_release();

	return next;
}

static proc_t *sched_next(hart_t hart, uint64_t *timeout)
{
	bool swapped = false;

	// Lock because other processes may be accessing the schedule
	lock_acquire(false);
	uint64_t rtc_slot = sched_rtc_slot();
	uint64_t offset = curr[hart] % MAX_TIME_SLOT;
	// Advance curr if the current slot has expired
	if (curr[hart] + schedule[hart][offset].length <= rtc_slot) {
		curr[hart] += schedule[hart][offset].length;
		swapped = true;
	}
	frame_t slot = schedule[hart][curr[hart] % MAX_TIME_SLOT];
	*timeout = slot2time(curr[hart] + slot.length);
	lock_release();
	// Release lock because we do not want to block when executing temporal fence.

	if (swapped) {
		temporal_fence(); // Insert a temporal fence if we swapped slots
	}

	if (slot.pid == INVALID_PID) {
		return NULL; // No process scheduled for this slot
	}

	// We now have a process we may schedule.
	proc_t *proc = proc_get(slot.pid);
	if (proc->timeout > slot2time(rtc_slot)) {
		return NULL; // Process is sleeping or waiting
	}

	// Try to acquire the process
	lock_acquire(false);
	bool proc_acquired = proc_acquire(slot.pid);
	proc->timeout = *timeout;
	lock_release();
	return proc_acquired ? proc : NULL;
}
\end{lstlisting}

That is, when a system call is made the TTAS lock is acquired, and when a new
task is scheduled the same lock is also acquired. Now while the preemptable
flag is true for the syscall, it checks that no preemption is pending while
gathering the lock, but once it has the lock, it will delay scheduling if the
scheduler runs while the syscall holds the lock. The syscalls in S3K attempt to
rectify this by checking for preemption after a constant amount of work so as
not to block the scheduler for too long, but if this constant time varies
between syscalls, it could be used in unintended ways and a domain could in
theory cause predictable scheduling delays and relay information to the next
domain. This analysis is purely theoretical and would require access to
hardware to fully test, and as such should mainly be seen as one of the
multitude of challenges a truly time protected kernel will face.

\subsection{The FreeRTOS kernel is vast}\label{sub:FreeRTOS-vast}
One requirement for time protection specified within \cite{ge2019} is that the
OS must be partitioned. This includes all system calls, which must be constant
time in regard to available information within each domain. Given that the
domain scheduler implemented creates a copy of the priority based scheduler for
each domain, it can be assumed that a large portion of these system calls
indirectly become constant time in this aspect. Further, as the user mode tasks
are unable to disable any interrupts this would indicate that the shared front
that remains to be partitioned is anything regarding the domain-level scheduler
itself. Given that these do make use of constant-time operations, and domain creation is
restricted to Privileged mode, then it would seem that this requirement might
be met to some extent. However, to truly show this partitioned OS one would
need a full audit of all system calls available for user mode tasks, and make
sure that they are constant time with respect to the domain available
information. As this is something beyond the scope of this thesis, the
implementation can provide no guarantees that it is truly implementing time
protection.

\subsection{Remaining timing channels and shared hardware}\label{sub:remaining-channels}
While scheduling timing channels are a significant vulnerability, they are far
from the only timing channel present in modern kernels. As such, the provided
port of the FreeRTOS kernel is by no means time protected. One significant form
of timing channel still available within the kernel is that of interrupts.
During development it was assumed that the only interrupt is that of the
machine timer used for scheduling. In a real system this is not the case. Here,
one would have to handle the external interrupts in a deterministic manner,
without breaking the scheduling guarantees of the domain-level scheduler. It
might also be desired that some domains are not interrupted during execution,
as this gives information about the availability of some external service that
in itself might be deemed an information leak. It is up to future work to
determine a method for handling these external interrupts in a manner where
they uphold both of the aforementioned issues.

Beyond the purely asynchronous software events a substantial amount of
remaining timing channels originate from non-isolated, shared hardware
structures. While this thesis makes use of the \texttt{fence.t} instruction to
clear the microarchitectural state on a domain switch, a lot of the hardware
still remains shared. Larger caches such as the L2 and L3 cache are too large
to simply flush \cite{wistoff2023}. This means higher level caches need another
method for isolation. In \cite{ge2019} they make use of cache coloring to
partition the higher level caches. In essence cache coloring works by using
some of the bits in the physical address to encode a color of the page. For
example, if using 2 bits in the physical address, one can encode 4 colors, and
when the application asks to allocate consecutive pages, then the OS makes sure
to map them to the correct colors \cite{dinh, WhatCacheColoring}. The L1 cache
is usually indexed directly with virtual addresses, and as such can not be
colored in this manner, whereas higher-level caches more often make use of
elements of the physical address.

Cache coloring utilises the cache in an unintended manner. In general, it
reduces the average performance of the entire system, while for some specific
applications it can improve performance by reducing cache interference.
However, in truly secure kernel software partitioning is still required and as
such it is a cost that is inevitable. The larger issue with cache coloring is
the uncertainty that it poses. Where the ISA usually standardizes the
architectural interface for virtual memory and the MMU behaviour, the same
cannot be said for cache coloring. For example, in RISC-V the ISA specifies the
programmer-visible MPU model, while the underlying cache implementation remains
undefined. This means that the cache coloring implementation is CPU-specific
based on things like cache indexing. This makes a cross platform implementation
of cache coloring challenging, and any future development in cache architecture
could break any such implementation \cite{WhatCacheColoring}.

In recent years different architectures have added support for hardware cache
partitioning. For example, Intel have added CAT
\cite{IntroductionCacheAllocation}, ARM with the MPAM extension
\cite{LearnarchitectureMemory} and similar work for RISC-V with the QoS
register interfaces \cite{11IntroductionRISCV}. For a truly secure kernel,
these items need to reach a maturity level, where domains can be guaranteed
separate in the higher level caches that previously could not be partitioned.

Another approach for safe OS execution is what was seen in S3K
\cite{karlsson2025}. Here they make use of the Scratch Pad Memory (SPM) which
is a section of memory that is never backed up to the higher level caches. The
kernel itself was then developed, such that all memory accesses by the kernel
only left a footprint in the SPM, which was then flushed on a domain switch.
This approach seems promising for the kernel data itself, assuming that the SPM
is guaranteed large enough to hold the kernel footprint. It might also offer
performance improvements over cache partitioning, as it could reduce the
interference the OS would have on user level applications. As such, it could be
a part of the solution, but it would still need partitioning of higher level
caches to ensure that user level application can remain partitioned.

Most likely there will never be a hardware solution which fits all. A general
purpose operating system, where user-level constant-time programming can solve
a large amount of issues, might not need specific domain separation and cache
partitioning. On the other hand, in high security environments this is a must
and something where depending on the use case the correct solution could vary.
It is up to future work to gather the understanding on where this boundary is,
and what hardware changes are needed for time protection.
