\section{Introduction}
Information channels are methods of communication in an unintended manner. One
such example is timing channels, which occur when a task's execution time
varies based on the execution of another task in the system. Whether
inadvertently or purposely, the other task is given unintended information
about the current task by altering its execution time.  In this way, timing
channels bypass the operating system's (OS's) security enforcement leaking
information through timing of observable events. At first many attacks were
targeting different cryptographic libraries and leaking information about
secret keys through these timing channels \cite{dhem2000, aciicmez2005}.
Further, attacks have shown that it is possible to gather information across
VMs running in a cloud computing setting \cite{ristenpart2009}. Not only that,
side-effects from out-of-order executions on modern processors gave way to the
Meltdown attacks to read arbitrary kernel-memory locations including personal
data and passwords \cite{lipp2018}. Different mitigations are moving forward,
such as hiding kernel memory from user space tasks \cite{corbet2017} to prevent
Meltdown attacks, and a move to more constant time programming in cryptographic
libraries with help from domain specific languages(DSL) such as FaCT
\cite{cauligi2019} that aims to guarantee constant-time execution with respect
to secretly marked information. While these mitigations do prove valuable,
modern CPUs' speculative execution still leaves doors open for more
sophisticated attacks such as the ones seen in \cite{kocher2019}.

The attack front is clear, but the solution is not as simple. While some
proposed solutions focus on the user side implementing secure programs, this
creates a huge burden on the developer to implement secure code. Instead
implementations of secure microkernels \cite{karlsson2025, ge2019} attempt to
guarantee security through the OS. The OS would be in charge of partitioning
domains, such that no channel between domains is possible. This is already done
for memory, and examples of formally proven memory-safe kernels are available
\cite{murray2013}. Attempts to create similar proof against the presence of
timing channels have been carried out \cite{buckley2023}, but as far as this
paper is aware no kernel is formally verified against timing channels.

Many of these attempts are focused on the creation of a highly secure OS from
the ground up, providing time protection as a foundation of the OS. While this
is a more robust solution for a truly safe OS the authors of \cite{ge2019} hint
that "there is no reason time protection cannot be implemented in other
systems...", something which could bring time protection to more mainstream
applications. However, adding time protection into an existing kernel is a
fundamentally different task than designing it from scratch.

FreeRTOS serves as the target platform for investigating this challenge.
FreeRTOS is a widely deployed real-time operating system, which relies on a
priority-based preemptive scheduler that is inherently vulnerable to
scheduling-based timing channels. While ground-up microkernels have
demonstrated that time protection is achievable, whether comparable mechanisms
can be integrated into an existing mainstream OS remains an open issue. The
thesis therefore addresses the following research question: \emph{Can time
protection mechanisms be integrated into a mainstream OS to mitigate timing
channels?}

To this end, this thesis investigates a defence mechanism composed of a
domain-level scheduler with deterministic flushing on domain switching via the
temporal fence instruction. This approach focuses primarily on mitigating
scheduling-based timing channels and leaks through the shared
microarchitectural state, both of which are discussed further in
Section~\ref{sub:timing_channels}.

To achieve this mechanism, this thesis adds the following contributions.
\begin{itemize}
  \item A domain-level scheduler that partitions execution into fixed time
    slices, ensuring execution of tasks in one domain does not affect the
    schedulability of a task in another domain.
  \item Temporal isolation with the use of temporal fence instruction during
    domain switches to clear shared microarchitectural state.
  \item An evaluation of the domain-level scheduler using QEMU system emulation
    with instruction-based time counting, showing constant-time domain
    scheduling.
\end{itemize}

The remainder of this thesis is organized as follows. Section~\ref{sec:Background}
provides the technical background on timing channels, the \texttt{fence.t}
instruction, and the FreeRTOS kernel. Section~\ref{sec:Related Work} surveys
related work on user-level mitigations, the seL4 time-protection kernel, and the
S3K real-time operating system. Section~\ref{sec:Scheduling} analyses the
FreeRTOS scheduler's vulnerability to timing channels and explores different
methods for closing them, and introduces the domain-level scheduler.
Section~\ref{sec:Implementation} details the implementation of this scheduler
within the FreeRTOS-MPU RISC-V port. Section~\ref{sec:Evaluation} presents the
evaluation using QEMU system emulation with two workloads and shows a constant
round trip using the cycle-count. Section~\ref{sec:FutureWork} discusses
directions for future work and the broader implications of partitioned hardware
and time protection. Finally, Section~\ref{sec:Conclusion} concludes the thesis.
